---
title: "R_notebook"
output: pdf_document
---

```{r}
#import librarys
library('dplyr')
library('data.table')
library('ggplot2')
library('reshape2')
library('RColorBrewer')
library('ggrepel')
library('lubridate')
library('chron')
library('janitor')
library('purrr')
```

```{r}
#read variables
airports <- read.csv("airports.csv")
carriers <- read.csv("carriers.csv")
airplanes <- read.csv("plane-data.csv")

#read data
df1 <- read.csv("2006.csv")
df2 <- read.csv("2007.csv")
df3 <- read.csv("2008.csv")

#datas without cancelled and diverted flights.

df1 <- df1 %>% filter(Cancelled == 0 & Diverted == 0)
df2 <- df2 %>% filter(Cancelled == 0 & Diverted == 0)

#explore data
df1
df2
```

```{r}
#dataset sizes 
nrow(df1)
nrow(df2)
```

```{r}
#sample to match size and avoid potential bias towards 2007
df2 <- sample_n(df2, 7003802)
nrow(df2) #verify
```
```{r}
# join dataframes
df <- df1 %>% full_join(df2) #join

#create delay column
df$delay <- (df$ArrDelay+df$DepDelay)
```

```{r}
#create usefull data and time columns

#date column
df$date<-as.Date(with(df,paste(Year,Month,DayofMonth,DayOfWeek, sep="-")),"%Y-%m-%d")

#time column
df$time_bin <- round((df$DepTime/100), digits= 0)
df$time <-times(df$DepTime / (100))
df <- df %>% 
          mutate(date_time = ymd_hm(paste(date, time)))

#create column with date bins
df$date_bin <- as.POSIXct(strptime(df$date_time, format = '%Y-%m-%d %H'))

df #explore 

```
```{r}
#look for missing values (DepDelay and ArrDelay should be removed due to cancelled&diverted = 0, reference = python_notebook)
colSums(is.na(df))
summary(df)
```

Questions

Question 1.1: 
When is the best day of the week to fly to minimise delays?

For mean:
```{r}
#map months and DayofWeek to corresponding names.
df$Month_name <- month.abb[df$Month] #month names
df$DayOfWeek <- recode(df$DayOfWeek, 
       "1"="Monday",
       "2"="Tuesday",
       "3"="Wednesday",
       "4"="Thursday",
       "5"="Friday",
       "6"="Saturday",
       "7"="Sunday") #weekday names

#create table df
day_mean <- df[c('DayOfWeek','DepDelay','ArrDelay')] #select columns
day_mean <- day_mean %>%
  group_by(DayOfWeek) %>%
  summarize(Dep_del = mean(DepDelay), Arr_del = mean(ArrDelay)) #group and average

#create columns in new df
day_mean$avg_delay_day <- ((day_mean$Dep_del+day_mean$Arr_del)/2)
day_mean$total_avg_delay_day <- ((day_mean$Dep_del+day_mean$Arr_del))
day_mean %>% 
group_by(total_avg_delay_day) %>% 
  arrange((total_avg_delay_day)) #group and arrange

```
Plots for mean:
```{r}
#prep data:
day_mean_plot <- day_mean %>%
  select(DayOfWeek, avg_delay_day)%>%
  group_by(DayOfWeek) %>% 
  arrange((avg_delay_day)) #group and arrange
day_mean_plot2 <- melt(day_mean, id.vars='DayOfWeek') #melt for pot 2 and 3

#plot 1
ggplot(day_mean_plot, aes(x=DayOfWeek, y=avg_delay_day, fill=DayOfWeek, color=DayOfWeek)) +
        geom_bar(stat='identity') +
        ylab("avg. delay(min.)") + xlab("Day of Week") +   
        labs(title = "avg. delay of wekdays") +
        theme(axis.text.x=element_text(angle=35,hjust=1))
#plot2
ggplot(day_mean_plot2, aes(x=DayOfWeek, y=value, fill=variable)) + 
        geom_bar(stat='identity') +
        ylab("distr. minutes") +
        theme(axis.text.x=element_text(angle=35,hjust=1))
#plot3
ggplot(day_mean_plot2, aes(x=DayOfWeek, y=value, fill=variable)) + 
       geom_bar(stat='identity', position='dodge') +
       ylab("Minutes") +
       theme(axis.text.x=element_text(angle=35,hjust=1))
```

For total and delay%:
```{r}
#create table df for flights
day_tot <- df[c('DayOfWeek','delay')] #select columns
day_tot <- day_tot %>%
  group_by(DayOfWeek) %>%
  tally %>% arrange(desc(n)) #total arrival delays
names(day_tot)[names(day_tot) == 'n'] <- 'Flights'
#create columns in day_tot df
day_tot$Flights_perc <- ((day_tot$Flights/14007604)*100)

#create table df for delays
day_tot2 <- df[c('DayOfWeek','DepDelay','ArrDelay')] #select columns
day_tot2 <- subset(day_tot2, DepDelay > 0 | ArrDelay > 0)
names(day_tot2)[names(day_tot2) == 'DepDelay'] <- 'Delay_perc'
day_tot2 <- day_tot2 %>%
  group_by(DayOfWeek) %>%
  tally %>% arrange(desc(n)) #total arrival delays
names(day_tot2)[names(day_tot2) == 'n'] <- 'Delays'

#join
day_df = left_join(day_tot, day_tot2, by = c("DayOfWeek" = "DayOfWeek"))
day_df$Delays_perc= (day_df$Delays/day_df$Flights) #create delay in percentage column

day_df #explore
```

Plots for total and delay%:
```{r}
#prep data:
day_df_plot <- day_df %>%
  select(DayOfWeek, Delays_perc)%>%
  group_by(DayOfWeek) %>% 
  arrange((Delays_perc)) #group and arrange

#plot 
ggplot(day_df_plot, aes(x=DayOfWeek, y=Delays_perc, fill=DayOfWeek, color=DayOfWeek)) +
        geom_bar(stat='identity') +
        ylab("Percentage") + xlab("Day of Week") +     
        labs(title = "avg. delay of wekdays (in %)") +
        theme(axis.text.x=element_text(angle=35,hjust=1))
```
The best day to fly to avoid delays is on Saturday or Tuesdays depending on tickets price.


Question 1.2
When is the best time of day to fly to minimise delays?
```{r}
#first we create time of day bin categories
categories<-c('night','before noon','afternoon','evening')
df$arr_time_bin<-cut(df$ArrTime, seq(0,2400,600),labels=categories)
df$dep_time_bin<-cut(df$DepTime, seq(0,2400,600),labels=categories)

#create time of day df
time_day = df[c('DepDelay','dep_time_bin','ArrDelay','arr_time_bin')]
time_day
```

```{r}
#Drop the missing values we found above.
time_day <- na.omit(time_day)
```

```{r}
#create total df's
arr_tot = time_day  %>% group_by(arr_time_bin) %>% 
tally %>% arrange(desc(n)) #total arrival delays
arr_tot #show

dep_tot = time_day  %>% group_by(dep_time_bin) %>% 
tally %>% arrange(desc(n)) #total departure delays
dep_tot #show
```
The total no. of delays is lowest at night (approx. 70.0000 delays.) But before concluding night is the best time: ets also find the mean delays:

```{r}
#create df's for mean
time_arr_mean <- time_day %>%
  group_by(arr_time_bin) %>%
  summarize(arr_mean = mean(ArrDelay))
time_arr_mean #show table

time_dep_mean <- time_day %>%
  group_by(dep_time_bin) %>%
  summarize(dep_mean = mean(DepDelay))
time_dep_mean #show table
```

```{r}
#summarize in one table

#time of day column with arr_time_bin & dep_time_bin as variables
names(time_arr_mean)[names(time_arr_mean) == 'arr_time_bin'] <- 'time_of_day'
names(time_dep_mean)[names(time_dep_mean) == 'dep_time_bin'] <- 'time_of_day'
dt_arr <- data.table(time_arr_mean, key = "time_of_day") 
dt_dep <- data.table(time_dep_mean, key = "time_of_day")

#mean
time_dep_mean <- dt_arr[dt_dep]
time_dep_mean$avg_delay_day <- ((time_dep_mean$dep_mean+time_dep_mean$arr_mean)) #avg. del column
time_dep_mean$total_delay_day <- ((time_dep_mean$dep_mean+time_dep_mean$arr_mean)) #total avg. del column

#group arrange
time_day_mean <- time_dep_mean %>% 
group_by(avg_delay_day) %>% 
  arrange((avg_delay_day)) 
time_day_mean
```
plot average delay in minutes:
```{r}
#prep data:
time_day_mean_plot <- time_day_mean %>%
  select(time_of_day, avg_delay_day)%>%
  arrange((avg_delay_day)) #group and arrange

#plot 
ggplot(time_day_mean_plot, aes(x=time_of_day, y=avg_delay_day, fill=time_of_day)) +
        geom_bar(stat='identity') +
        ylab("Percentage") + xlab("Day of Week") +     
        labs(title = "Percentage of delays by time of day") +
        theme(axis.text.x=element_text(angle=35,hjust=1))
```

Lets also have a look at flights, delay, and delay%
```{r}
#create total dep columns
time_day_dep <- time_day %>% group_by(dep_time_bin) %>% 
tally %>% arrange(desc(n))
names(time_day_dep)[names(time_day_dep) == 'n'] <- 'Flights'

#create total arr columns
time_day_arr <- time_day %>% group_by(arr_time_bin) %>% 
tally %>% arrange(desc(n))
names(time_day_arr)[names(time_day_arr) == 'n'] <- 'Flights'

#join
time_day_tot = left_join(time_day_dep, time_day_arr, by = c("dep_time_bin" = "arr_time_bin"))
names(time_day_tot)[names(time_day_tot) == 'dep_time_bin'] <- 'TimeofDay'
time_day_tot$Flights <- time_day_tot$Flights.x+time_day_tot$Flights.y
time_day_tot<- time_day_tot %>% select('TimeofDay', 'Flights')

#df dep delays columns
time_day_dep2 <- subset(time_day, DepDelay > 0)
time_day_dep2 <- time_day_dep2 %>%
  group_by(dep_time_bin) %>%
  tally %>% arrange(desc(n)) #total arrival delays
names(time_day_dep2)[names(time_day_dep2) == 'n'] <- 'Delays'

#df arr delays columns
time_day_arr2 <- subset(time_day, ArrDelay > 0)
time_day_arr2 <- time_day_arr2 %>%
  group_by(arr_time_bin) %>%
  tally %>% arrange(desc(n)) #total arrival delays
names(time_day_arr2)[names(time_day_arr2) == 'n'] <- 'Delays'

#join
time_day_tot2 = left_join(time_day_dep2, time_day_arr2, by = c("dep_time_bin" = "arr_time_bin"))
names(time_day_tot2)[names(time_day_tot2) == 'dep_time_bin'] <- 'TimeofDay'
time_day_tot2$Delays <- time_day_tot2$Delays.x+time_day_tot2$Delays.y
time_day_tot2<- time_day_tot2 %>% select('TimeofDay', 'Delays')

#final join and show table
time_day_tot3 = left_join(time_day_tot, time_day_tot2, by = c("TimeofDay" = "TimeofDay"))
time_day_tot3$Delays_perc= (time_day_tot3$Delays/time_day_tot3$Flights) #create delay in percentage column
time_day_tot3
```

Plot percentage of delays by time of day:
```{r}
#prep data:
time_day_tot3_plot <- time_day_tot3 %>%
  select(TimeofDay, Delays_perc)%>%
  arrange((Delays_perc)) #group and arrange

#plot 
ggplot(time_day_tot3_plot, aes(x=TimeofDay, y=Delays_perc, fill=TimeofDay)) +
        geom_bar(stat='identity') +
        ylab("Percentage") + xlab("Day of Week") +     
        labs(title = "Percentage of delays by time of day") +
        theme(axis.text.x=element_text(angle=35,hjust=1))
```

The best time to travel to avoid any delays are before noon. Interesting, we can see that our first observation (lowest total number of delays at night) doesn't mean that traveling at night is ultimately the best time as it comes with the risk of much longer delays, if they occur.

Question 1.3
When is the best time of year to fly to minimise delays?
```{r}
#for mean:

#create df with columns of interest
month_mean <- df[c('Month','DepDelay','ArrDelay')]
month_mean <- month_mean %>%
  group_by(Month) %>%
  summarize(Dep_del = mean(DepDelay), Arr_del = mean(ArrDelay)) #group and avg.

#average columns
month_mean$avg_delay_day <- ((month_mean$Dep_del+month_mean$Arr_del)/2) #avg. column
month_mean$tot_avg_delay_day <- (month_mean$Dep_del+month_mean$Arr_del) #tot column

#group and arrange
month_mean %>% 
group_by(avg_delay_day) %>% 
  arrange((avg_delay_day)) 
```
```{r}
#total and delay%

#create df for flight total
month_tot <- df[c('Month_name','delay')]
month_tot <- month_tot %>%
  group_by(Month_name) %>%
  tally %>% arrange(desc(n)) 
names(month_tot)[names(month_tot) == 'n'] <- 'Flights'
month_tot$Flights_perc <-  ((month_tot$Flights/14007604)*100)
month_tot

#create df for delays
month_tot2 <- df[c('Month_name','DepDelay','ArrDelay')]
month_tot2 <- subset(month_tot2, DepDelay > 0 | ArrDelay > 0)
month_tot2 <- month_tot2 %>%
  group_by(Month_name) %>%
  tally %>% arrange(desc(n)) 
names(month_tot2)[names(month_tot2) == 'n'] <- 'Delays'
#month_tot$Delays_perc <-  ((month_tot2$delay/14007604)*100)
month_tot2

#join
month_tot3 = left_join(month_tot, month_tot2, by = c("Month_name" = "Month_name"))
month_tot3$Delays_perc = (month_tot3$Delays/month_tot3$Flights) #create delay in percentage column
month_tot3 <- arrange(month_tot3,Delays_perc) #arrange
month_tot3
```
```{r}
#delays by month piechart
pie(month_tot3$Delays, month_tot3$Month_name)
```
The best time of year to travel to avoid delays are in September. The christmass and summer holidays are the worst. If we wanted to choose by season, we can do the following:
```{r}
#choose months and add
winter <- month_mean[12, ] + month_mean[1, ]+ month_mean[2, ]
spring <- month_mean[3, ] + month_mean[4, ]+ month_mean[5, ]
summer <- month_mean[6, ] + month_mean[7, ]+ month_mean[8, ]
autumn <- month_mean[9, ] + month_mean[10, ]+ month_mean[11, ]

#print results
print(paste("Total average delay during the spring: ", spring$tot_avg_delay_day))
print(paste("Total average delay during the summer: ", summer$tot_avg_delay_day))
print(paste("Total average delay during the autumn: ", autumn$tot_avg_delay_day))
print(paste("Total average delay during the winter: ", winter$tot_avg_delay_day))
```
Autumn is the season that best avoids delays:

Question 2.
Do older planes suffer more delays?
```{r}
#join df and airplane df
names(df)[names(df) == 'TailNum'] <- 'tailnum' #match name of variable df airplane
#join
df_airplanes <- inner_join(df,
  select(airplanes, tailnum, year),
  by = "tailnum"
)

#edit the joined df
names(df_airplanes)[names(df_airplanes) == 'year'] <- 'plane_year' #change col name
df_airplanes$plane_year <- as.numeric(df_airplanes$plane_year) #change plane_year column to numeric
df_airplanes <- df_airplanes[!is.na(df_airplanes$plane_year),] #Drop NA's

df_airplanes #explore 
```
We define a plane old as +20 years (1987) (ref: ttps://www.paramountbusinessjets.com/faq/age-of-aircraft-safety-factor.html)

```{r}
#add plane_condition column and define category
df_airplanes <- df_airplanes %>% 
  mutate(plane_condition = if_else(plane_year <= 1987, "old", "new"))

#total no. of old and new planes
df_airplanes %>% group_by(plane_condition) %>% 
tally %>% arrange(desc(n)) #explore with conditions
```
There is significant more new planes, so lets look at the mean.  
```{r}
#mean delay for old vs. new planes
plane_mean <- df_airplanes %>%
  group_by(plane_condition) %>%
  summarize(delay = mean(delay))
names(plane_mean)[names(plane_mean) == 'delay'] <- 'Total_avg_delay'

plane_mean #explore
```

```{r}
#total for plane condition

#create df for flight total
plane_tot <- df_airplanes[c('plane_condition','delay')]
plane_tot <- plane_tot %>%
  group_by(plane_condition) %>%
  tally %>% arrange(desc(n)) 
names(plane_tot)[names(plane_tot) == 'n'] <- 'Flights'
plane_tot$Flights_perc <- (plane_tot$Flights/(10100221+1232687)*100)
plane_tot

#create df for delays
plane_tot2 <- df_airplanes[c('plane_condition','DepDelay','ArrDelay')]
plane_tot2 <- subset(plane_tot2, DepDelay > 0 | ArrDelay >0)
plane_tot2 <- plane_tot2 %>%
  group_by(plane_condition) %>%
  tally %>% arrange(desc(n)) 
names(plane_tot2)[names(plane_tot2) == 'n'] <- 'Delays'
plane_tot2

#join
plane_tot3 = left_join(plane_tot, plane_tot2, by = c("plane_condition" = "plane_condition"))
plane_tot3$Delays_perc = (plane_tot3$Delays/plane_tot3$Flights) #create delay in percentage column
plane_tot3 <- arrange(plane_tot3,Delays_perc) #arrange

#join with mean
plane_all = left_join(plane_tot3, plane_mean, by = c("plane_condition" = "plane_condition"))
plane_all #explore

```
plots:
```{r}
#plot 1
ggplot(plane_mean, aes(x=plane_condition, y=Total_avg_delay, fill=plane_condition, color=plane_condition)) +
        geom_bar(stat='identity') +
        ylab("Minutes") + xlab("Plane condition") +     
        labs(title = "Mean time of delays for planes") +
        theme(axis.text.x=element_text(angle=35,hjust=1))
#plot2
ggplot(plane_all, aes(x=plane_condition, y=Delays_perc, fill=plane_condition, color=plane_condition)) +
        geom_bar(stat='identity') +
        ylab("Percentage") + xlab("Plane condition") +     
        labs(title = "Percentage of delayed flights") +
        theme(axis.text.x=element_text(angle=35,hjust=1))
```


```{r}
100-(20.44398/20.61757)*100
```
The mean delay of old vs new planes is almost identical. New planes only have 0.84% decrease in mean delay. Hence, older does not suffer significant more delays.

Try changing the threshold age to 10 year:
```{r}
#Create age categories
categories<-c('1956-1987(old)','1987-1997(mid)','1997-2007(new)') #categories
df_airplanes$plane_condition2<-cut(df_airplanes$plane_year, seq(1977,2007,10),labels=categories) #sequence
df_airplanes$plane_condition2[is.na(df_airplanes$plane_condition2)] = "1956-1987(old)" #set <1987 to old
#group and arrange
df_airplanes %>% group_by(plane_condition2) %>% 
tally %>% arrange(desc(n)) #explore
```

```{r}
#again, we take the mean delay for old/mid/new planes 
plane2_mean <- df_airplanes %>%
  group_by(plane_condition2) %>%
  summarize(delay = mean(delay)) #explore the mean
names(plane2_mean)[names(plane2_mean) == 'delay'] <- 'Total_avg_delay'

plane2_mean
```

```{r}
#total for plane condition in 3 bins : old, new, mid

#create df for flight total
plane2_tot <- df_airplanes[c('plane_condition2','delay')]
plane2_tot <- plane2_tot %>%
  group_by(plane_condition2) %>%
  tally %>% arrange(desc(n)) 
names(plane2_tot)[names(plane2_tot) == 'n'] <- 'Flights'
plane2_tot$Flights_perc <- (plane2_tot$Flights/(10100221+1232687)*100)
plane2_tot

#create df for delays
plane2_tot2 <- df_airplanes[c('plane_condition2','DepDelay','ArrDelay')]
plane2_tot2 <- subset(plane2_tot2, DepDelay > 0 | ArrDelay >0)
plane2_tot2 <- plane2_tot2 %>%
  group_by(plane_condition2) %>%
  tally %>% arrange(desc(n)) 
names(plane2_tot2)[names(plane2_tot2) == 'n'] <- 'Delays'
plane2_tot2

#join
plane2_tot3 = left_join(plane2_tot, plane2_tot2, by = c("plane_condition2" = "plane_condition2"))
plane2_tot3$Delays_perc = (plane2_tot3$Delays/plane2_tot3$Flights) #create delay in percentage column
plane2_tot3 <- arrange(plane2_tot3,Delays_perc) #arrange

#join with mean
plane2_all = left_join(plane2_tot3, plane2_mean, by = c("plane_condition2" = "plane_condition2"))

plane2_all
```
Plot for old, mid, new bins:
```{r}
#plot 1
ggplot(plane2_mean, aes(x=plane_condition2, y=Total_avg_delay, fill=plane_condition2, color=plane_condition2)) +
        geom_bar(stat='identity') +
        ylab("Minutes") + xlab("Plane condition") +     
        labs(title = "Mean time of delays for planes") +
        theme(axis.text.x=element_text(angle=35,hjust=1))
#plot2
ggplot(plane2_all, aes(x=plane_condition2, y=Delays_perc, fill=plane_condition2, color=plane_condition2)) +
        geom_bar(stat='identity') +
        ylab("Percentage") + xlab("Plane condition") +     
        labs(title = "Percentage of delayed flights") +
        theme(axis.text.x=element_text(angle=35,hjust=1))
```
Old planes does not suffer significantly more delay than newer planes. Although there is an interesting increase in delay around the 10 year mark.

Question 3.
How does the number of people flying between different locations change over time?
```{r}
#load data for 11 years with selected columns and count flights. 11 years seems more reasonable for finding changes, rather 
#than only the initial 2 years used untill now.
df_97 = read.csv("1997.csv")[ ,c("Origin", "Dest", "Year")] %>%
  group_by(Origin, Dest) %>%
  summarise(y1997 = n()) %>%
  arrange(desc(y1997)) %>%
  collect()

df_98 = read.csv("1998.csv")[ ,c("Origin", "Dest", "Year")] %>%
  group_by(Origin, Dest) %>%
  summarise(y1998 = n()) %>%
  arrange(desc(y1998)) %>%
  collect()

df_99 = read.csv("1999.csv")[ ,c("Origin", "Dest", "Year")] %>%
  group_by(Origin, Dest) %>%
  summarise(y1999 = n()) %>%
  arrange(desc(y1999)) %>%
  collect()

df_00 = read.csv("2000.csv")[ ,c("Origin", "Dest", "Year")] %>%
  group_by(Origin, Dest) %>%
  summarise(y2000 = n()) %>%
  arrange(desc(y2000)) %>%
  collect()

df_01 = read.csv("2001.csv")[ ,c("Origin", "Dest", "Year")] %>%
  group_by(Origin, Dest) %>%
  summarise(y2001 = n()) %>%
  arrange(desc(y2001)) %>%
  collect()

df_02 = read.csv("2002.csv")[ ,c("Origin", "Dest", "Year")] %>%
  group_by(Origin, Dest) %>%
  summarise(y2002 = n()) %>%
  arrange(desc(y2002)) %>%
  collect()

df_03 = read.csv("2003.csv")[ ,c("Origin", "Dest", "Year")] %>%
  group_by(Origin, Dest) %>%
  summarise(y2003 = n()) %>%
  arrange(desc(y2003)) %>%
  collect

df_04 = read.csv("2004.csv")[ ,c("Origin", "Dest", "Year")] %>%
  group_by(Origin, Dest) %>%
  summarise(y2004 = n()) %>%
  arrange(desc(y2004)) %>%
  collect()

df_05 = read.csv("2005.csv")[ ,c("Origin", "Dest", "Year")] %>%
  group_by(Origin, Dest) %>%
  summarise(y2005 = n()) %>%
  arrange(desc(y2005)) %>%
  collect()

df_06 = read.csv("2006.csv")[ ,c("Origin", "Dest", "Year")] %>%
  group_by(Origin, Dest) %>%
  summarise(y2006 = n()) %>%
  arrange(desc(y2006)) %>%
  collect()

df_07 = read.csv("2007.csv")[ ,c("Origin", "Dest", "Year")] %>%
  group_by(Origin, Dest) %>%
  summarise(y2007 = n()) %>%
  arrange(desc(y2007)) %>%
  collect()
```

```{r}
#merge all the dataframes
dataframes <- list(df_97, df_98,df_99, df_00,df_01, df_02,df_03, df_04,df_05, df_06, df_07)
df_locations <- purrr::reduce(.x = dataframes, merge, by = c("Origin", "Dest"), all = T)
head(df_locations,10)
```

```{r}
#clean the new df

#replace nan with 0, = no flights
df_locations <- df_locations %>%
  mutate_all(funs(ifelse(is.na(.), 0, .))) 

df_locations$Total <- rowSums(df_locations[,3:13] ) #create new total col
df_locations <- df_locations %>% arrange(desc(Total)) #arrange
df_locations <- head(df_locations,20) #select only top 20 
df_locations
```

```{r}
#add locations together:
con1 <- df_locations[1,3:14] + df_locations[2,3:14]
con1$connection = c("LAS&LAX")
con1 <- con1 %>%
  select(connection, everything())

con2 <- df_locations[3,3:14] + df_locations[4,3:14]
con2$connection = c("PHX&LAX")
con2 <- con2 %>%
  select(connection, everything())

con3 <- df_locations[5,3:14] + df_locations[6,3:14]
con3$connection = c("SFO&LAX")
con3 <- con3 %>%
  select(connection, everything())

con4  <- df_locations[7,3:14] + df_locations[8,3:14]
con4$connection = c("ORD&MSP")
con4 <- con4 %>%
  select(connection, everything())

con5  <- df_locations[9,3:14] + df_locations[10,3:14]
con5$connection = c("PHX&LAS")
con5 <- con5 %>%
  select(connection, everything())

con6  <- df_locations[11,3:14] + df_locations[12,3:14]
con6$connection = c("LGA&ORD")
con6 <- con6 %>%
  select(connection, everything())

con7  <- df_locations[13,3:14] + df_locations[19,3:14]
con7$connection = c("HOU&DAL")
co7n <- con7 %>%
  select(connection, everything())

con8  <- df_locations[14,3:14] + df_locations[15,3:14]
con8$connection = c("OAK&LAX")
con8 <- con8 %>%
  select(connection, everything())

con9  <- df_locations[16,3:14] + df_locations[20,3:14]
con9$connection = c("EWR&ORD")
con9 <- con9 %>%
  select(connection, everything())

con10  <- df_locations[17,3:14] + df_locations[18,3:14]
con10$connection = c("BOS&LGA")
con10 <- con10 %>%
  select(connection, everything())

dataframes2 <- list(con1,con2,con3,con4,con5,con6,con7,con8,con9,con10) #select frames
df_locations2 <- purrr::reduce(.x = dataframes2, merge, all = T) #merge 

df_locations2 <- df_locations2 %>% arrange(desc(Total)) #arrange
df_locations2 #explore
```

```{r}
# add total by year row
library(janitor)
df_locations3 <- df_locations2 %>%
  adorn_totals("row")
df_locations3
```
We can see that people flying between locations have generally fallen from 1997 to 2007, with a noticeable decrease in 2001 due to 9/11.

plot a map over connections
```{r}
#create df's needed for a map plot

#count connections
connections <- df_locations %>%
  group_by(Origin, Dest,Total) %>%
  summarise(n = sum(Total)) %>%
  arrange(desc(Total)) %>%
  collect()

#manuel rank
rank_list= c(1,1,2,2,3,3,4,4,5,5,6,6,7,7,8,8,9,9,10,10)
connections$rank <-rank_list
connections

#2nd df with lat and long data from airports ds for line between airports
connections2 = left_join(connections, airports, by = c("Origin" = "iata")) %>%
  select(Origin, Dest, n, Origin_Lat = lat, Origin_Long = long) %>%
  left_join(airports, by = c("Dest" = "iata")) %>%
  select(Origin, Dest, n, Origin_Lat, Origin_Long, Dest_Lat = lat, Dest_Long = long)

#3nd df for the labels of airports
connections3_1 = connections %>%
  group_by(Origin) %>%
  summarise(n = sum(n)) %>%
  arrange(desc(n))
#join
connections3 = left_join(connections3_1, airports, by = c("Origin" = "iata")) %>%
  select(Origin, n, Origin_Lat = lat, Origin_Long = long)
conections3 <- within(connections3, rm(n)) #remove n column
```
```{r}
#plot the maps with df created above

connections %>%
  filter() %>%
  group_by(Origin, Dest) %>%
  #summarise(n = sum(n, na.rm = TRUE)) %>%
  inner_join(airports, by = c("Dest" = "iata")) %>%
  ggplot(aes(y = lat, x = long)) +
  borders("state", fill = "lightgrey") +
  geom_point(aes(colour = rank), alpha = 0.4, size = 7) + 
  ggtitle("Map of US flights between locations (1997-2007)") +theme(plot.title = element_text(hjust = 0.5)) +
  geom_segment(data = connections2,
              mapping = aes(x = Origin_Long, y = Origin_Lat, xend = Dest_Long, yend = Dest_Lat),
              alpha = 1, colour = "orange", size = 0.8) +
  geom_label_repel(aes(x = Origin_Long, y = Origin_Lat, label = Origin), data = connections3) +
  coord_map(xlim = c(-125, -65), ylim = c(25, 50))+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position="none")
  #ggsave("Q3_1.png")
```

Let's explore the data further with more plots:
```{r}
#data prep
df_locations2_plot <- df_locations2[,1:12] #plot 1 data
df_locations2_plot2 <- df_locations2 %>%
  select(connection,y1997, y2001, y2002, y2007) #plot 2 data
df_locations2_plot3 <- df_locations3[11,1:12] #plot 3 data

#plot 4: 2001 exploration
df_01_2 = read.csv("2001.csv")[ ,c("Origin", "Dest", "Month")] %>%
  group_by(Month) %>%
  summarise(y2001 = n()) %>%
  arrange(Month) %>%
  collect()
df_01_2$Month_name <- month.abb[df_01_2$Month]

#melt 
df_locate <- melt(df_locations2_plot, id.vars='connection') #melt 1 data
df_locate2 <- melt(df_locations2_plot2, id.vars='connection') #melt 2 data
df_locate3 <- melt(df_locations2_plot3, id.vars='connection') #melt 3 data


ggplot(df_locate, aes(x=connection, y=value, fill=variable)) + 
        geom_bar(stat='identity') +
        ylab("Count") +
        theme(axis.text.x=element_text(angle=35,hjust=1))

ggplot(df_locate2, aes(x=connection, y=value, fill=variable)) + 
       geom_bar(stat='identity', position='dodge') +
       ylab("Count") +
       theme(axis.text.x=element_text(angle=35,hjust=1))

ggplot(df_locate3, aes(x=variable, y=value, group= 1)) + 
       geom_line(color = 4, size = 1.3) +
       ggtitle("Total no. of flights between locations (1997-2007)") +
       xlab("Year") +
       ylab("Total no. of flights") +
       theme_minimal()+ 
       theme(
       plot.title=element_text( hjust=0.5, vjust=0.5)) 
#ggsave("Q3_2.png", bg = "#FFFFFF")

ggplot(df_01_2, aes(x=Month_name, y=y2001, group= 1)) + 
       geom_line(color = 4, size = 1.3) + scale_x_discrete(limits = month.abb)+
       ggtitle("Exploration of drop in 2001") +
       xlab("Month") +
       ylab("Total no. of flights") +
       theme_minimal()+ 
       theme(
       plot.title=element_text( hjust=0.5, vjust=0.5)) 

```

Question 4.
Can you detect cascading failures as delays in one airport create delays in others?
```{r}
#create df's for two airports las vegas and los angeles

df_airport <- df

#create las vegas df
df_LAS <- subset(df_airport, Origin == "LAS" & Dest =="LAX")

df_LAS <- df_LAS %>%
  select(Origin,DepDelay,Dest,DepTime,ArrDelay,date_bin,date_time)

df_LAS$DepDelay <-as.numeric(df_LAS$DepDelay > 0)
df_LAS$ArrDelay <-as.numeric(df_LAS$ArrDelay > 0)

df_LAS <- subset(df_LAS, ArrDelay == 1)
df_LAS <- arrange(df_LAS, date_time)

input_time <- df_LAS$date_bin

df_LAS$date_bin <-  input_time + lubridate::hours(1) # add 1 hour to bin (due to the flight is around 1h, so we need the impact on the later bin for next df)
df_LAS$date_bin <- as.character(df_LAS$date_bin)
df_LAS$date_time <- as.character(df_LAS$date_time)
df_LAS$DepDelay <- as.integer(df_LAS$DepDelay)

#group the df by date 
df_LAS_group <- df_LAS %>%
  group_by(date_bin) %>%
  summarize(delay_in_A = sum(ArrDelay))

df_LAS
df_LAS_group
```

```{r}
#create Los Angeles to all other airports
df_LAX <- subset(df_airport, Origin == "LAX" & Dest =="LAS")

df_LAX <- df_LAX %>%
  select(Origin,DepDelay,Dest,ArrDelay,DepTime,date_bin, date_time)

df_LAX$DepDelay <-as.numeric(df_LAX$DepDelay > 0)
df_LAX$ArrDelay <-as.numeric(df_LAX$ArrDelay > 0)

df_LAX <- subset(df_LAX, DepDelay == 1 & ArrDelay == 1)
df_LAX <- arrange(df_LAX, date_time)

df_LAX$date_bin <- as.character(df_LAX$date_bin)
df_LAX$date_time <- as.character(df_LAX$date_time)
df_LAX$DepDelay <- as.integer(df_LAX$DepDelay)

df_LAX_group <- df_LAX %>%
  group_by(date_bin) %>%
  summarize(delay_in_B_and_C = sum(DepDelay, ArrDelay)/2)

df_LAX
df_LAX_group
```
```{r}
#join on date bin
airport_A_B_A <- inner_join(df_LAX_group, df_LAS_group, by= "date_bin")
airport_A_B_A
```

```{r}
#select only where delay in airport B&C is bigger or equal to airport A delay!
airport_A_B_A <- subset(airport_A_B_A, delay_in_B_and_C >= delay_in_A)
colSums(airport_A_B_A[,3]) #count no of cascading failures
airport_A_B_A 
```

```{r}
#cascading delay percentage of total flights on the route
3659/26014*100
```
We can detect 3661 cascading delay failures for airport A (LAS) that creates delays in airport B (LAX) and C (LAS again), approx 14.1% of all the flights leads to cascading delays. 


Let's also try for delay in LAS(A), that create delay in LAX(B) and then in all other airports(C).
```{r}
#create Los Angeles to all other airports
df_LAX_all <- subset(df_airport, Origin == "LAX")

df_LAX_all <- df_LAX_all %>%
  select(Origin,DepDelay,Dest,ArrDelay,DepTime,date_bin, date_time)

df_LAX_all$DepDelay <-as.numeric(df_LAX_all$DepDelay > 0)
df_LAX_all$ArrDelay <-as.numeric(df_LAX_all$ArrDelay > 0)

df_LAX_all <- subset(df_LAX_all, DepDelay == 1 & ArrDelay == 1)
df_LAX_all <- arrange(df_LAX_all, date_time)

df_LAX_all$date_bin <- as.character(df_LAX_all$date_bin)
df_LAX_all$date_time <- as.character(df_LAX_all$date_time)
df_LAX_all$DepDelay <- as.integer(df_LAX_all$DepDelay)

df_LAX_all_group <- df_LAX_all %>%
  group_by(date_bin) %>%
  summarize(delay_in_B_and_C = sum(DepDelay, ArrDelay)/2)

df_LAX_all
df_LAX_all_group
```

```{r}
#join on date bin
airport_A_B_C <- inner_join(df_LAX_all_group, df_LAS_group, by= "date_bin")
airport_A_B_C
```
```{r}
#select only where delay_in_a is smaller, there should be less filtered out this time according to the data above.
airport_A_B_C <- subset(airport_A_B_C, delay_in_B_and_C >= delay_in_A)
colSums(airport_A_B_C[,3]) #count no of cascading failures
airport_A_B_C 
```
```{r}
#tot percentage of the flights from A->B->C
12866/455560*100
```
We can detect 12894 cascading failures for airport A (LAS) that creates delays in airport B & C (LAX and all others), approx 2.83% of all the flights leads to cascading delays. 


Question 5.
Use the available variables to construct a model that predicts delays
```{r}
#create df for ml (we use df1 and df2 where cancelled and diverted are excluded)
df_ml <- df1 %>% full_join(df2) #join dfs for ml

#prepare for model
df_ml$delay <- (df_ml$ArrDelay+df_ml$DepDelay) #create delay column
df_ml$delay_binary <- ifelse(df_ml$delay>0, 1, 0) #create target column in binary. 0 = on time, 1 = delay.
df_ml <- df_ml[, which(names(df_ml) != "delay")] #drop delay column
df_ml <- sample_n(df_ml, 250000) #sample due to computational limitations
df_ml$delay_binary <- as.factor(df_ml$delay_binary) # convert target column to factor

df_ml #explore
```

```{r}
#load random forest library (we know from the python notebook that it performs well)
library(randomForest)

#splitting into training and test set(80/20)
set.seed(100)
ml_training_set <- sample(nrow(df_ml), 0.8*nrow(df_ml), replace = FALSE)
X <- df_ml[ml_training_set,]
Y <- df_ml[-ml_training_set,]

#create model
ml_model <- randomForest(delay_binary ~ ., data = X, importance = TRUE)
ml_model


#predictions
pred <- predict(ml_model, Y, type = "class") #classification predictions
pred <- mean(pred == Y$delay_binary) #target mean
print(paste("Random Forrest model prediction : ",pred))

#plots

#class error over amount of trees
#png('Q5_1.png') # create png
plot(ml_model,lwd=1, main="Class error over no. of trees")
#dev.off() #save png

#roc curve
library(ROCR) #load roc curve library for prediction and plotting
#png('Q5_2.png') # create png
pred2 = predict(ml_model,type = "prob") #predict with probability
pred2 = prediction(pred2[,2], X$delay_bin)
pred2 = performance(pred2, 'tpr','fpr') #use performance function from ROCR library
plot(pred2,lwd=2,col=4,main="Binary Random Forest model in R") #set linewidth, color, title and  plot.
#dev.off() #save png

```











